---
title: "Global average half-life per cell type"
output:
  html_notebook:
    toc: true
    number_sections: true
    theme: readable
author: Marat Mufteev
date: "`r format(Sys.time(), 'Generated on %d %B, %Y')`"
---

Upload libraries
```{r}
library(dplyr)
library(ggplot2)
```

Upload half-life (HL) data from a saturation curve fit
```{r}
HL_tbl = read.csv("./HL_saturation_curve.csv.gz",
                  header = T)
```

To measure average half-life (HL) we attempted multiple methods.
Genes included for the calculation:
A) all genes with measured HL
B) unreliable genes are removed (degree of stringency could be varied)
C) HL values for unreliable genes are interpolated from a ratio method
We noticed that the conclusion on a global HL shift didn't depend on the method.
```{r}
# Method B 
# Thresholds are somewhat arbitrary
# However, the higher average HL in RTT neurons is not too sensitive to the thresholds in a reasonable range
HL_clean = HL_tbl %>%
           group_by(celltype) %>%
           filter(TR > quantile(TR, 0.1, na.rm = T) | 
                  SS > quantile(SS, 0.1, na.rm = T),
                  CI50_HL < 0.75) %>%
           ungroup() %>%
           select(gene, celltype, HL)
```

Visualize a global shift in HL
```{r}
ggplot(data = HL_clean,
       aes(x = log(HL, 10),
           color = celltype)) +
  stat_density(geom = "line", position = "identity", size = 1) +
  scale_color_manual(values = c("#FFAA00", "#0092FF")) +
  ggthemes::theme_base(base_size = 25) +
  scale_x_continuous(breaks = c(0, 0.4771213, log(9, 10)), labels = c(1, 3, 9), 
                     limits = c(log(0.5, 10), log(24, 10))) +
  theme(axis.title.y=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks.y=element_blank()) +
  theme(plot.background = element_rect(fill = "white",colour = "white",size = 1))
```






